Global land characterisation using land cover fractions at 100 m resolution

### 2020-11-18 article/article.tex
Draft 2 of first review round

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
%\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
\usepackage{textcomp} % for degree sign
%\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
\usepackage{graphicx}
%\usepackage{adjustbox} % For resizing tables
\usepackage{array} % For mid-alignment (m{xcm}) in tables
\usepackage{rotating} % Sideways figures
%\bibliographystyle{apalike} % elsarticle defined at the top

@@

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}
% Package names
\newcommand{\cran}[1]{CRAN: #1}

@@

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SIF}{SIF}{sun-induced chlorophyll fluorescence}
\newacronym{GPP}{GPP}{gross primary productivity}
\newacronym{MSI}{MSI}{multi-spectral instrument}
\newacronym{SAR}{SAR}{synthetic aperture radar}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{classification and regression tree}
\newacronym{GLM}{GLM}{general linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
\newacronym{GDAL}{GDAL}{Geospatial Data Abstraction Library}
%\makenoidxglossaries

@@

\title{Global land characterisation using cover fractions at 100m resolution}

@@

\begin{abstract} % elsarticle: environment instead of a command \abstract{
Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a predetermined ratio. In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity. To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.

We compared the performance of three types of machine learning regression algorithms for land cover fraction mapping.
For the best model, we assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe at a sampling interval of 0.2 degrees.
The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points of the \gls{CGLS-LC100} project. Both datasets are global and aligned with the PROBA-V 100 m grid.

One challenge for models for land cover fraction mapping is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well. We proposed a new solution by combining three models. A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.

Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%. Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved by a tuned random forest model with our proposed three-model approach. The models were driven primarily by optical remote sensing data, followed by soil and climate data. Herbaceous cover was the most challenging to discern for all models, whereas built-up and water were the easiest to estimate.

This research proves that machine learning algorithms can be applied globally to a wide variety of land cover classes to create an exhaustive global land cover fraction map. Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which helps to customise the result for a diverse range of uses. In addition, this work directly contributes to the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project, as well as to the accuracy assessment of land cover fraction maps both thematically and spatially.
\end{abstract}

@@

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received increasing amounts of attention, also as new satellites are launched and the capacity for land monitoring increases.
In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} map \citep{esa_land_2017} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

Except for the \ac{CGLS-LC100} product with its cover fraction layers, all other global land cover maps, including previous maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
This issue becomes worse at coarse resolutions and heterogeneous landscapes.
It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus underestimate tree cover in the pixel.
These systematic errors then add up when scaling the result to the entire region.

A potential solution to this issue is to characterise land cover using cover fractions.
In this approach, instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
That way, the land cover models work not on pixel labels, such as ``forest'', but on land cover characteristics, such as tree cover, defined as the area of the pixel covered by tree canopies, as well as herbaceous cover, defined as the area not covered by woody vegetation.
This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but these concepts include the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.

Land cover fraction mapping has been attempted in the past.
Most of the previous research has focused on deriving land cover fractions of 3-6 classes at a local scale  \citep{adams_classification_1995, foody1996fuzzyevaluation, walton2008subpixelrf, hansen_continuous_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018}, less often at a regional scale and with more detailed classes \citep{colditz_land_2011}.
The methods for assessing the accuracy of the results vary greatly between the different studies.
In addition, global land cover fraction products have emerged, but focused on a particular class, such as tree cover \citep{hansen_global_2003, Hansen2013forestchange}, water \citep{schroeder_development_2015} and urban area \citep{gao_mapping_2020}.
To date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression \citep{walton2008subpixelrf}, \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
So far other studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.

In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}. In addition to optical covariates (such as blue, red, \gls{NIR} and shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}. Elevation covariates (elevation, slope, aspect) are known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}. \Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are often used to improve land cover class separability, e.g. for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}. Therefore in this paper such covariates have also been used in order to increase the accuracy of land cover fraction regression model predictions, and their importance for the predictions was analysed.

A common issue with the use of land cover fraction data as input into regression models is data imbalance.
The more classes are mapped, the more often there will be 0\% of a given class in any particular pixel, leading to zero inflation.
In addition, in homogeneous areas, 100\% cover of a single class is common.
This leaves little data in the middle for the regression models to learn from.
Conversely, regression models tend to favour predictions closer to the mean and rarely predict extreme values.
We investigated this issue and suggested ways to deal with it.

Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
It requires a comprehensive dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
In this study, we made use of an independent validation set collected as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019} that includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as in our study.
The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
We used the validation data to assess model accuracy both thematically and spatially.

All in all, in this study we investigated and compared methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
We also investigated ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
Therefore the objectives of our study were to:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping.
 \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean.
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions.
 \item Assess the accuracy of the models from a thematic and spatial point of view.
\end{enumerate}

@@

The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020, nandin-erdene_tsendbazar_copernicus_2019}.
The training set was generated by experts in \ac{IIASA}.
This global point dataset provides information on fractions of each class within sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
The fraction estimates were acquired by high-resolution image interpretation by experts, for the year 2015.
Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.

For model accuracy assessment, an independent global validation dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.

The data includes over 150\,000 training points and over 21\,000 validation points across the globe, describing the fractions of 12 classes in the year 2015.
The classification scheme follows the \gls{LCCS} \citep{buchhorn_copernicus_2020}.
However, due to limited number of observations for some rare classes, we merged them to get a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous vegetation (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
The unknown class was discarded: points with the dominant land cover class marked as unknown were not used.
For points with a minority fraction of unknown, the remaining classes were linearly rescaled to add up to 100\%.
Thus in the end the training dataset size became 138\,164 and the validation dataset size became 20\,705. See figure \ref{fig-reference-data} for the spatial distribution of the points.

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/maps/2020-07-06-training-and-validation}
 \caption{Sample points representing 100x100 m areas at which land cover reference data was used in the study. The colours represent the dominant land cover class at each point. (a): training dataset, collected by \gls{IIASA}, 138\,164 points used in this study. (b): validation dataset, collected by \gls{WUR}, 20\,705 points used in this study. Both datasets were collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020}.}
 \label{fig-reference-data}
\end{figure}

\subsection{Model covariates}

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
See figure \ref{fig-processing} for an overview of the whole processing chain.
The processing was carried out in R \citep{r_2019} and the resulting code has been made openly available in \citet{dainius_masiliunas_2020_3973123}.

In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
These covariates had to be preprocessed before they could be input into the models.

@@

% \begin{table}
% \centering
%     \begin{tabular}{llp{6cm}}
%          \toprule
%          \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
%          \midrule
%          Spectral & 22 & PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
%          Harmonic & 9 & PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
%          Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
%          Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
%          Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
%          Location & 3 & None (intrinsic) \\
%          \bottomrule
%     \end{tabular}
%     \label{tab-inputdata}
%     \caption{Data sources for the covariates input into the model.}
% \end{table}

@@

Moreover, we produced boxplots showing the distribution of predictions, in bins for each 10\% of the predictions, against the validation data distribution, and a comparison with the 1:1 line.
Lastly, we produced spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point.
We rasterised the result into a map for ease of view, taking the mean value where several points fell into the same raster cell.

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
%\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
\usepackage{textcomp} % for degree sign
%\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
\usepackage{graphicx}
\usepackage{adjustbox} % For resizing tables
\usepackage{array} % For mid-alignment (m{xcm}) in tables
\usepackage{rotating} % Sideways figures
%\bibliographystyle{apalike} % elsarticle defined at the top

@@

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}
% Package names
\newcommand{\cran}[1]{CRAN:~#1}

@@

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{RRMSE}{RRMSE}{relative root mean squared error}
\newacronym{RMAE}{RMAE}{relative mean absolute error}
\newacronym{RME}{RME}{relative mean error}
\newacronym{NSE}{NSE}{the Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{the International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SIF}{SIF}{sun-induced chlorophyll fluorescence}
\newacronym{GPP}{GPP}{gross primary productivity}
\newacronym{MSI}{MSI}{multi-spectral instrument}
\newacronym{SAR}{SAR}{synthetic aperture radar}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{classification and regression tree}
\newacronym{GLM}{GLM}{general linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
\newacronym{GDAL}{GDAL}{the Geospatial Data Abstraction Library}
\newacronym{GFCC}{GFCC}{Global Forest Cover Change}
\newacronym{GHSL}{GHSL}{Global Human Settlement Layer}
\newacronym{GSW}{GSW}{Global Surface Water}
%\makenoidxglossaries

@@

\title{Global land characterisation using land cover fractions at 100 m resolution}

@@

\begin{abstract} % elsarticle: environment instead of a command \abstract{
Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a predetermined ratio. In contrast, land cover fraction mapping enables expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity. To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.

We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping.
For the model with the lowest MAE, we assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points, provided by the \gls{CGLS-LC100} project. Both datasets are global and aligned with the PROBA-V 100 m UTM grid.

A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% fractions are difficult to predict accurately. We proposed a new solution by combining three models. A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.

Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%. Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest with our proposed three-model approach. This model was driven primarily by optical remote sensing data, followed by climate data.

This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions. Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which enables customising the result for a diverse range of uses. The three-step approach enables mapping 0\% and 100\% fractions more accurately, and thus has already been taken up in the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project. This study also includes an accuracy assessment of land cover fraction maps both thematically and spatially, which could be taken up by future land cover fraction mapping efforts.
\end{abstract}

@@

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has lately received more attention due to increased availability of higher spatial and temporal resolution satellite data.
In this context, the capacity for land monitoring has increased, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} product \citep{esa_land_2017} that provides a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} product \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that provides a finer spatial resolution and higher quality with yearly updates since 2015, and the \ac{FROM-GLC} product \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

Except for the cover fraction layers of the \ac{CGLS-LC100} product, all other global land cover products that include all major land cover classes, such as the ones described by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single land cover class.
Such discrete classification oversimplifies reality, as mixed pixels that are covered by multiple land cover classes are a common occurrence. 
This issue is exacerbated at coarse resolutions and over heterogeneous landscapes.
It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus underestimate tree cover in the pixel.
These systematic errors then add up when scaling the result to the entire region.

A potential solution to this issue is to characterise land cover using cover fractions.
In this approach, instead of a single discrete class, the proportion of every class in the legend is reported for every pixel of the map.
That way, the land cover models work not on pixel labels, such as ``forest'', but on land cover characteristics, such as tree cover, defined as the area of the pixel covered by tree canopies, or herbaceous cover, defined as the area not covered by woody vegetation.
This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' mapping or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
%A related term is ``super-resolution mapping'', but this concept includes the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.

Land cover fraction mapping has been attempted in the past.
Most of the previous research has focused on deriving land cover fractions of 3-6 classes at a local scale  \citep{adams_classification_1995, foody1996fuzzyevaluation, walton2008subpixelrf, hansen_continuous_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018}, less often at a regional scale and with more detailed classes \citep{colditz_land_2011}.
The methods for assessing the accuracy of the results vary greatly between the different studies.
In addition, global land cover fraction products have emerged, but focused on a particular class, such as tree cover \citep{hansen_global_2003, Hansen2013forestchange, townshend_global_2017}, water \citep{schroeder_development_2015, pekel_high-resolution_2016} and urban area \citep{corbane_automated_2019, gao_mapping_2020, gong_annual_2020}.
To date, only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Methods that have been tested in previous studies include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression (e.g. \citealp{walton2008subpixelrf}), \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
The previous studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.

A common issue with the use of land cover fraction data as input into regression models is data imbalance.
The more classes are mapped, the more more likely it is that one or more classes are not present in a given pixel (have a 0\% fraction), leading to zero inflation.
This is especially the case in homogeneous areas, where we can find not only an inflation of 0\% fractions, but also an inflation in 100\% fractions.
This leaves little data in the middle for the regression models to learn from.
Conversely, regression models tend to favour predictions closer to the mean and rarely predict extreme values.
In this study, we propose the use of a hierarchical multi-step modelling approach to better predict these extreme values.

In addition to classification methods, one aspect that is important for classification accuracy is the data (features) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
In addition to optical features (such as spectral bands), temporal features (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are commonly used to improve land cover class separability, e.g. for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
Elevation features (elevation, slope, aspect) are also known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
%In this paper, we also used such features for increasing the accuracy of land cover fraction regression model predictions.
%, and their importance for the predictions was analysed.

Lastly, model accuracy assessment by itself is often challenging, especially at a global scale.
It requires a comprehensive dataset across the globe that would be comparable with the training data, and yet independent of it.
Accuracy assessment of land cover fractions is even more challenging, as it requires a dataset that either provides fraction information, or fine spatial resolution data from which it can be calculated.
Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited.
This makes it difficult for the users to decide whether a given product suits their needs.
Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.

%In this study, we made use of a validation dataset that was collected independently from the training dataset, as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019}.
%The dataset includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as the training dataset.
%The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
%We used the validation data to assess model accuracy both thematically and spatially.

In this study we investigated and compared a variety of methods for performing global land cover fraction mapping.
Furthermore, we proposed a method to enhance land cover fraction mapping further by dealing with data imbalance, which is inherent in land cover fraction training data, and tested it on the method that achieved the highest accuracy in the method comparison.
Lastly, we investigated the effect of a wide variety of input features on each mapped class fraction, to gain insight into what data was key for the model with the highest accuracy to achieve the best result.
Therefore the objectives of our study were to:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning regression models for mapping fractions of land cover on a global scale.
 \item Investigate approaches for reducing bias in the model predictions with regards to zero inflation and predictions tending towards the mean.
 \item Analyse the importance of features input into the highest accuracy model for the prediction of each land cover fraction.
 \item Assess the accuracy of the models from a thematic and spatial point of view, comparing it to existing global land cover products.
\end{enumerate}

@@

The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020, nandin-erdene_tsendbazar_copernicus_2019}.
The data includes over 150\,000 training points and over 21\,000 validation points across the globe, describing the fractions of 12 classes in the year 2015.
The classification scheme follows the \gls{LCCS} \citep{buchhorn_copernicus_2020}.
However, due to the limited number of observations for some rare classes, we merged them to get a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous vegetation (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
The ``unknown'' class was discarded: points with the dominant land cover class marked as unknown were not used.
For points with a minority fraction of unknown, the remaining classes were linearly rescaled to add up to 100\%.
Thus in the end the training dataset size became 138\,164 and the validation dataset size became 20\,705. See figure \ref{fig-reference-data} for the spatial distribution of the points.

These global datasets were generated by performing high-resolution satellite imagery interpretation by a team of experts, using the GeoWiki platform.
Each sampled point corresponded to a single 100~m by 100~m pixel of the PROBA-V 100~m UTM grid \citep{buchhorn_copernicus_2020}.
The area of each of these sampled pixels was subdivided into 100 subpixels at 10~m by 10~m spatial resolution.
The subpixels were labelled by the experts, and then converted into land cover fraction estimates by calculating the proportion of subpixels that each land cover class covers in the pixel.
The satellite imagery that the experts interpreted corresponded to the year 2015.

%These global point datasets provide information on fractions of each class within sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
%The fraction estimates were acquired by high-resolution image interpretation by experts, for the year 2015.
%Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.

The training set was generated by a team at \ac{IIASA}, whereas the validation dataset was generated by a team at \gls{WUR}.
The validation dataset was first developed over Africa \citep{tsendbazar_developing_2018}, and was later expanded to cover the whole world \citep{buchhorn_copernicus_2020}.
The class definitions and tools used to collect the data was equivalent for both datasets, but performed independently by a separate group of regional experts to ensure independence of the data, and using a different sampling method.
The validation dataset was generated separately using stratified random sampling and follows the CEOS Land Product Validation guidelines, which focus on independent and statistically rigorous accuracy assessment.
In contrast, the training dataset uses a mix of systematic sampling and additional random sampling in areas that exhibit higher heterogeneity, so as to provide the algorithms with enough data to train in all areas of the world.

\begin{figure}
 \includegraphics[width=\textwidth]{article-figures/maps/2020-10-12-training-and-validation}
 \caption{Sample points representing 100x100~m areas at which land cover reference data was used in the study. The colours represent the dominant land cover class at each point. (a): training dataset, collected by \gls{IIASA}, 138\,164 points used in this study. (b): validation dataset, collected by \gls{WUR}, 20\,705 points used in this study. Both datasets were collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020}.}
 \label{fig-reference-data}
\end{figure}

\subsection{Model training features}

See figure \ref{fig-processing} for an overview of the whole processing chain used in this study.
The processing was carried out in R \citep{r_2019} and the resulting code has been made openly available in \citet{dainius_masiliunas_2020_3973123}.

To train the models and predict land cover fractions in unsampled locations, six groups of features were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
These features had to be preprocessed before they could be input into the models.

@@

% \begin{table}
% \centering
%     \begin{tabular}{llp{6cm}}
%          \toprule
%          \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
%          \midrule
%          Spectral & 22 & PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
%          Harmonic & 9 & PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
%          Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
%          Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
%          Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
%          Location & 3 & None (intrinsic) \\
%          \bottomrule
%     \end{tabular}
%     \label{tab-inputdata}
%     \caption{Data sources for the features input into the model.}
% \end{table}

@@

Moreover, to show the variation in predicted land cover fraction values depending on the magnitude of the fractions, we produced boxplots showing the distribution of the predictions.
The boxplots are binned for each 10\% of the predictions.
The 1:1 line indicates how well do the distributions of the predictions and the reference data match.

Lastly, we produced spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point.
We rasterised the result into a map for ease of view.
If multiple points fell into the same raster cell, we reported the mean value.
